Solar Flares Across the Spectrum: Observations and Instrumentation

### 2025-04-11 chapter2/chap2.tex
Local changes

The remarkable technological progress attained in the last few decades has yielded significant benefits in the form of highly advanced imaging, spectroscopic, and polarimetric instruments designed for astronomical observations. These instruments have empowered us with the ability to scrutinize the Sun with exceptional detail. Specifically, the space based observatories have added new avenues of observing sun. Large portions of the electromagnetic spectra is heavily by Earth's atmosphere \sr{Reference here}.

The studies within this thesis are using solar observations are multi wavelength analysis of solar flares with both imaging and spectroscopy. In this chapter we describe the details of the instruments and techniques used for our analysis.

@@

The Solar Dynamics Observatory \citep[SDO;][]{sdo} is a NASA mission designed to understand the causes of solar variability at various spatio temporal and wavelength scales, as well as its impacts on Earth and the near-Earth environment. It is part of NASA’s Living With a Star program and houses multiple instruments onboard to address its numerous scientific goals. Among the numerous instruments on-board {\it SDO}, we have extensively used data from two instruments – the Atmospheric Imaging Assembly (AIA) and the Helioseismic and Magnetic Imager (HMI).

@@

AIA \citep{aia} obtains full-disk images of the solar transition region and corona at unprecedented spatio-temporal resolution. The field of view (FOV) covers up to 0.5 $\mathrm{R_{\odot}}$ above the limb. It incorporates several filters at 94 {\AA}, 131 {\AA}, 171 {\AA}, 193 {\AA}, 211 {\AA} and 335 {\AA}, mainly probing the Corona, along with filters at 304 {\AA}, 1600 {\AA}, 1700 {\AA}, and 4500 {\AA}, mainly observing the Chromosphere and Photosphere, providing a total temperature coverage range of 0.06 MK to 20 MK and across various heights. Figure 2.1 illustrates the temperature responses of these filters \citep{borner12}, derived based on the CHIANTI model for solar emissivity \citep{chianti,chianti1}. Under different conditions, these AIA filters observe different plasma phenomena \citep{o'dwyer10}. Typical exposure times range between 0.5 and 3 seconds. Each AIA telescope is accompanied by its own guide telescope to ensure highly accurate image stabilization. For ous analysis, we have extensively used the coronal EUV channel observations from AIA, e.g. 94 {\AA}, 131 {\AA}, 171 {\AA}, 193 {\AA}, 211 {\AA} and 335 {\AA}. These coronal channels observations are usually available with a 12 s cadence.

@@

HMI \citep{hmi} obtains images of the Sun across the \ion{Fe}{1} 6173 {\AA} line at six wavelength locations and infers all four components of the Stokes' parameter. Using these observations, various parameters are inferred e.g., Line of Sight (LOS) magnetic field, Doppler velocities, vector magnetic fields etc. We have mainly used the LOS magnetogram for our analysis. These magnetograms have a 0.5{\arcsec} per pixel resolution and 45 s cadence.

@@

The Interface Region Imaging Spectrograph \citep[IRIS;][]{iris} is a NASA small satellite explorer mission designed to observe the dynamics of the lower solar atmosphere. IRIS contains a spectrograph and a slit-jaw imager (SJI), observing the chromosphere, the transition region, and the lower corona. It primarily observes the Sun in two pass bands around 1400 {\AA} and 2800 {\AA}. IRIS provides data at high spatial resolution (0.33 {\arcsec} in FUV and 0.4 {\arcsec} in NUV), high time cadence (up to 1 second), and high spectral resolution (12 or 25 m{\AA} per pixel).

IRIS observes three wavelength bands: two in the Far Ultraviolet (FUV) range, spanning 1331.7–1358.4 {\AA} and 1389.0–1407.0 {\AA}, and one in the Near Ultraviolet (NUV) range, covering 2782.7–2851.1 {\AA}. IRIS measures spectra through two primary modes: raster and sit-and-stare. In raster mode, the slit traverses a field of view, capturing spectra from each pixel along its path. When the displacement between consecutive slit positions is comparable to the slit width, it is called dense raster mode; otherwise, it is called sparse raster mode. Alternatively, the instrument may position the slit on a specific region for observation of the Sun. This mode can allow the Sun to naturally rotate across the region or adjust for solar rotation while maintaining continuous observation. This mode is referred to as sit-and-stare. In our analysis, we have used both spectra and SJI imaging from the \ion{Mg}{2} h \& k window and SJI imaging from the \ion{Si}{4} 1402 {\AA} window.

@@

Solar Orbiter \citep[SO;][]{so} is a joint mission between the European Space Agency (ESA) and NASA, launched in February 2020 with the goal of studying the Sun and its dynamic behaviour. The spacecraft aims to provide unprecedented observations of the Sun's polar regions and the inner heliosphere, offering new insights into solar phenomena such as solar wind, coronal mass ejections, and the Sun's magnetic field. We have used observations from the Spectrometer/Telescope for Imaging X-rays (STIX) instrument for our analysis.

 STIX \citep{stix,stix1} is designed to study the Sun's X-ray emissions, providing crucial insights into the processes driving solar flares and other high-energy phenomena. STIX incorporates a set of collimators and detectors to capture high-resolution images of the Sun in X-rays. These images allow scientists to pinpoint the locations and intensities of X-ray emissions associated with solar flares. The instrument covers a broad range of X-ray energies, from approximately 4 to 150 keV, enabling it to observe various aspects of solar activity. STIX can capture X-ray data with high temporal resolution, allowing scientists to study the rapid evolution of solar flares and other transient events in detail. STIX works together with other instruments onboard Solar Orbiter, providing complementary measurements to enhance our understanding of the Sun's dynamic behaviour across different wavelengths and energy ranges.

@@

Solar Terrestrial Relations Observatory-Ahead \citep[{\it STEREO-A};][]{stereo}, is one of two spacecraft in NASA's STEREO mission, which was launched in October 2006 with the goal of studying the Sun and its dynamic behaviour. {\it STEREO-A} and its twin spacecraft {\it STEREO-B} (Behind) were designed to provide stereoscopic views of the Sun, enabling three-dimensional observations of solar phenomena such as coronal mass ejections (CMEs) and solar flares. Extreme Ultraviolet Imager \citep[{\it EUVI};][]{euvi} is one of the key instruments onboard the {\it STEREO-A} spacecraft. {\it EUVI} is designed to capture images of the Sun in the extreme ultraviolet (EUV) wavelength range. This range of wavelengths is particularly useful for studying the Sun's outer atmosphere, known as the corona, and observing features such as solar flares, coronal loops, and coronal holes.

{\it EUVI} can capture high-resolution images of the Sun's corona, with spatial resolutions in order of a few arcseconds. It can observe the Sun in 171 {\AA}, 195 {\AA}, 284 {\AA} and 304 {\AA} simultaneously. {\it EUVI} provides continuous observations of the Sun from the STEREO-A spacecraft's vantage point, located slightly ahead of Earth in its orbit around the Sun. This allows {\it EUVI} to monitor solar activity over extended periods and to track the evolution of solar features such as sunspots, flares, and coronal mass ejections (CMEs). We have used 171 {\AA} and 195 {\AA} observations from EUVI for our analysis.

@@

{\it Hinode} \citep{hinode}, meaning "Sunrise" in Japanese, is a space mission launched by the Japan Aerospace Exploration Agency (JAXA) in collaboration with NASA and the UK Space Agency in September 2006. Also known as Solar-B before its launch, {\it Hinode} is dedicated to studying the Sun's magnetic field and its dynamic behaviour, focusing on understanding the mechanisms driving solar activity and space weather phenomena. The X-Ray Telescope\citep[XRT;][]{xrt} is one of the key instruments aboard the {\it Hinode} spacecraft.

{XRT captures high-resolution solar corona images in X-ray wavelengths, allowing us to study phenomena such as solar flares, coronal loops, etc. Its imaging capabilities enable us to investigate the solar corona's structure, dynamics, and heating mechanisms. XRT utilizes narrow-band filters to isolate specific X-ray emission lines emitted by highly ionized elements in the solar corona. XRT provides continuous observations of the solar corona from the vantage point of the {\it Hinode} spacecraft, which orbits the Earth in a Sun-synchronous polar orbit. This allows XRT to monitor solar activity over extended periods. In our analysis, we used XRT imaging observation from various filters.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Geostationary Operational Environmental Satellites (GOES)}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Geostationary Operational Environmental Satellites (GOES) are a series of weather satellites operated by the National Oceanic and Atmospheric Administration (NOAA) in the United States. These satellites continuously monitor weather conditions across the Americas, including the United States, Canada, Mexico, and parts of South America. GOES data are used by meteorologists and weather forecasters to monitor and predict weather conditions, track severe weather events, and issue warnings and advisories to the public. The continuous monitoring provided by GOES satellites helps improve the accuracy and timeliness of weather forecasts. In addition to monitoring terrestrial weather, GOES satellites also monitor space weather conditions, including solar flares, coronal mass ejections (CMEs), and geomagnetic storms. This information is crucial for forecasting space weather events and assessing their potential impact on satellite communications, navigation systems, and power grids. Among the numerous instruments on-board {\GOES}, we have used data from two instruments extensively {--} X-Ray Sensor (XRS) and Solar Ultra Violet Imager (SUVI).

@@

The X-Ray Sensor \citep[XRS;][]{xrs} is an instrument onboard {\it GOES}. The primary function of the XRS is to continuously monitor and measure full disk integrated solar soft X-ray emissions, particularly those associated with solar flares. XRS is designed to detect and measure solar X-ray emissions in two energy bands: short wavelength (0.5 to 4 {\AA}) and long wavelength (1 to 8 {\AA}).

@@

Solar Ultraviolet Imager \citep[SUVI;][]{suvi} is an instrument aboard {\it GOES}, designed to observe the Sun in the EUV spectrum, providing crucial data for monitoring solar activity and space weather. SUVI captures images of the Sun in 94 {\AA}, 131 {\AA}, 171 {\AA}, 195 {\AA}, 284 {\AA} and 304 {\AA}. These wavelengths are very similar to AIA and provide complementary observations. The FoV of SUVI is rotated by $\sim~45^{o}$, and hence, provides a larger radial distance coverage along the diagonals. In our analysis, we have used SUVI observations from 94 {\AA}, 131 {\AA}, 171 {\AA}, and 195 {\AA}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Various Plasma Diagnostic Methods}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We employ various techniques throughout this thesis to infer electron density, ion and electron temperature, velocity, thermal structure of the plasma observed in both Chromosphere and Corona. It is standard practice to use atomic databases like CHIANTI \citep{chianti,chianti1} to derive the plasma properties. Below we describe the working principle of the methods used within the scope of this thesis.

@@

Differential emission measure analysis is a commonly employed method with instruments that observe multiple spectral bands to determine the temperature and emission measure of optically thin coronal plasma exhibiting multiple thermal components along a line of sight. However, the inversion process is often challenging due to its ill-posed nature and the frequent lack of sufficient constraints, making it an under determined problem. One of the very first DEM calculation algorithm was {\it xrt\_dem\_iterative}, which was designed and validated for use with XRT data \citep{golub04,weber04}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Plasma Velocity from Spectra}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Numerous modern and historical imaging, spectroscopic, and polarimetric instruments have significantly contributed to our comprehension of the Sun and its atmosphere. Several of these instruments, previously discussed, have been utilized in this thesis. While certain instruments are ground-based telescopes, others operate from space. Space-based instruments primarily observe the Sun in Ultraviolet (UV), extreme ultraviolet (EUV), and X-ray bands, capturing radiation emitted from upper atmospheric layers like the transition region and corona, which possess elevated temperatures. Over recent decades, various studies of the Sun have effectively unveiled the physical characteristics of the gas in its upper atmospheric layers using observations recorded by X-ray imaging (e.g., {\it Hinode} X-ray Telescope \citep[{\it Hindoe}/XRT,][]{xrt}) and spectroscopy (e.g., the Spectrometer/Telescope for Imaging X-rays on {\it Solar Orbiter} \citep[{\it SO}/STIX,][]{stix}) as well as extreme ultraviolet (EUV) imaging (e.g., Atmospheric Imaging Assembly on the {\it Solar Dynamic Observatory} \cite[{\it SDO}/AIA,][]{aia}; Extreme ultraviolet Imaging Telescope on {\it Solar and Heliospheric Observatory} \cite[{\it SoHO}/EIT,][]{eit}; Solar Ultraviolet Imager on the {\it Geostationary Operational Environmental Satellites} \citep[{\it GOES}/SUVI,][]{suvi}; the Extreme Ultraviolet Imager on {\it Solar Terrestrial Relations Observatory-A} \cite[{\it STEREO-A}/EUVI,][]{stereo,euvi}; the Extreme-Ultraviolet Imager on {\it SO} \citep[{\it SO}/EUI][]{eui}) and spectroscopy (e.g., \citep[Hinode/EIS,][]{eis}), among others. Instruments such as the {\it Transition Region and Coronal Explorer} \citep[{\it TRACE},][]{trace}, the Solar Ultraviolet Measurements of Emitted Radiation on {\it SoHO} \citep[{\it SoHO}/SUMER,][]{sumer}, and {\it Interface Region Imaging Spectrograph} \citep[{\it IRIS},][]{iris} have facilitated detailed examinations of the chromosphere and the transition region. These missions offer continuous full-disk and Region of Interest (RoI) coverage of the Sun across X-ray and EUV wavelengths. However, the scenario is notably different in the Near-Ultraviolet (NUV) regime, where there is evidently a lack of continuous full solar disk coverage within this wavelength range.

A significant challenge in conducting solar observations in the UV range from the ground is the substantial attenuation caused by the Earth's atmosphere. To overcome this obstacle, one of the pioneering approaches involved the deployment of stratospheric balloon-borne instruments in 1970 and 1971, as documented by \cite{herse79}. This instrument featured a 20 cm telescope capable of imaging the Sun within the 200-460 nm range. Subsequently, the Rasolba balloon experiment utilized a 30 cm telescope equipped with an ultraviolet spectrograph, yielding high-resolution spectra of the Sun spanning 190 to 295 nm \citep{samain85,staath95}. Building upon these endeavors, the Sunrise project \citep{sunrise1,sunrise2} emerged as a balloon-borne observatory designed to observe the Sun in the Near Ultraviolet (NUV) range using a 1 m diameter telescope. Sunrise conducted two flights, in June 2009 and June 2013, respectively, and provided high-resolution imaging at wavelengths of 214, 300, 312, 388, and 397 nm with the Sunrise Filter Imager \citep[SuFI,][]{sufi} in 2009, and at 214, 279, and 397 nm in 2017 \citep{sunrise2}. Additionally, Dopplergrams and vector magnetograms were obtained in \ion{Fe}{1} 525.02 nm using the Imaging Magnetograph eXperiment \citep[IMaX,][]{imax} at various locations on the solar disk. Throughout these flights, Sunrise observed a variety of solar phenomena, including emerging flux events \citep{centeno17}, properties and dynamics of moving magnetic features around pores \citep{kaithakkal17}, proper motion of bright points in quiet sun and active regions \citep{jafarzadeh17}, and properties of fibrils \citep{gaferia17}. These observations underscored the wealth of information carried by this wavelength range and paved the way for full-disk coverage of the Sun in the NUV.

The Solar Ultraviolet Imaging Telescope \citep[SUIT;][]{ghosh16,article}, iss one of the seven payloads aboard the Aditya-L1 mission \citep{adityal1, aditya} led by the Indian Space Research Organization (ISRO). Launched on September 2, 2023, the satellite orbits around the Sun-Earth L1 point in a halo orbit. Equipped with eleven science filters, comprising eight narrow band and three broad band filters, SUIT possesses a distinctive capability to explore various heights within the solar photosphere and chromosphere. This capability aids in understanding the diverse physical processes involved in the transport of mass and energy across different layers of the Sun. Additionally, SUIT offers a unique opportunity to spatially measure the solar spectral irradiance in the Near Ultraviolet (NUV) range, a critical aspect in advancing our comprehension of Sun-climate relationships.

{\suit} is designed to deliver continuous full-disk and Region of Interest (RoI) solar images, boasting a plate scale of 0.7 \arcsec. It possesses the ability to track RoIs while compensating for the effects of differential rotation, as outlined by \cite{suit_algo}. Additionally, SUIT is equipped with onboard intelligence to detect and localize flares, and it can autonomously adjust exposure times to prevent saturation. The primary scientific inquiries targeted by SUIT include \citep{suit_science, suit_main}:

\begin{itemize}
    \item dynamic coupling between the lower and the middle solar atmosphere
    \item measurement of solar spectral radiance in the NUV.
    \item spectral energy distribution of solar flares in NUV
    \item dynamics of chromospheric eruptive phenomena at various spatio temporal scales
    \item Initiation of Coronal Mass Ejections(CMEs) and space weather: the kinematics of erupting prominence during the early phase.
\end{itemize}

@@

The Solar Ultraviolet Imaging Telescope \citep[SUIT;][]{ghosh16,article} is one of the seven payloads onboard the Aditya-L1 mission \citep{adityal1} of the Indian Space Research Organization (ISRO). With its 11 science filters (3 broadband and eight narrow band), SUIT will have the capability to probe different heights of solar atmosphere in photosphere and chromosphere to help us understand the various processes that transport mass and energy from one layer to another. SUIT will provide full disk as well as partial disk images of the Sun with a pixel size of 0.7". Through SUIT imaging, we would be able to resolve solar flares spatially on the surface of the Sun, for the first time in near ultra-violet (NUV), which will help us to address the questions regarding their build-up and triggering mechanism. In addition, to measure the spatially resolved solar spectral irradiance within the wavelength range that is central for studying the Chemistry of oxygen and ozone in the Stratosphere of Earth's atmosphere.

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[width = \linewidth]{Figures/Krucker_2015_ApJ_802_19_page-0004.jpg}
    \caption{X-ray and optical imaging of the three flares at the peak time of the impulsive phase: the images are from HMI with the pre-flare image subtracted. The contours represent RHESSI Clean maps in the thermal (red, 12–15 keV) and non-thermal (blue, 30–80 keV) HXR range \citep{krucker15}.}
    \label{fig1}
\end{figure}

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[width = 0.7\linewidth]{Figures/nphys1741_page-0002.png}
    \caption{Averaged TSI variations during flares. The TSI time-series averages over four exclusive sets of solar flares of decreasing amplitude. The black and pink curves correspond respectively to the TSI measured by the PMOV6 and the DIARAD radiometers. The dashed lines correspond to the 95\% confidence levels, while the vertical line denotes the peak time of the flare \citep{kretzschmar10}.}
    \label{fig2}
\end{figure}

>>>>

The remarkable technological progress attained in the last few decades has yielded significant benefits in the form of highly advanced imaging, spectroscopic, and polarimetric instruments designed for astronomical observations. These instruments have empowered us with the ability to scrutinize the Sun with exceptional detail. Specifically, the space based observatories have added new avenues of observing the Sun. The studies within this thesis are performed using multi-wavelength imaging and spectroscopic observations of solar flares recorded from space. This chapter describes the details of the instruments and techniques used for our analysis.

@@

The Solar Dynamics Observatory \citep[SDO;][]{sdo} is a NASA mission designed to understand the causes of solar variability at various spatio-temporal and wavelength scales, as well as its impacts on Earth and the near-Earth environment. It is part of NASA’s Living With a Star program and houses multiple instruments onboard to address its numerous scientific goals. Among the three instruments onboard {\it SDO}, we have extensively used data from two instruments – the Atmospheric Imaging Assembly (AIA) and the Helioseismic and Magnetic Imager (HMI).

@@

AIA \citep{aia} obtains full-disk images of the solar transition region and corona with 0.6{\arcsec} pixel size and 12s cadence in EUV channels and 24s cadence in UV channel. The field of view (FOV) covers up to 0.5 $\mathrm{R_{\odot}}$ above the limb. It incorporates several filters at 94~{\AA}, 131~{\AA}, 171~{\AA}, 193~{\AA}, 211~{\AA} and 335~{\AA}, mainly probing the corona, along with filters at 304 {\AA}, 1600 {\AA}, 1700 {\AA}, and 4500 {\AA}, mainly observing the chromosphere and photosphere, providing a temperature coverage range of 0.06 MK to 20 MK and across various heights. Under different conditions, these AIA filters observe different plasma phenomena \citep{o'dwyer10}. Typical exposure times range between 0.5 and 3 seconds. Each AIA telescope is accompanied by its own guide telescope to ensure highly accurate image stabilization. For our analysis, we have extensively used the coronal EUV channel observations from AIA, e.g. 94 {\AA}, 131 {\AA}, 171 {\AA}, 193 {\AA}, 211 {\AA} and 335 {\AA}. These coronal channels observations are usually available with a 12 s cadence.

@@

HMI \citep{hmi} obtains images of the Sun across the \ion{Fe}{1} 6173 {\AA} line at six wavelength locations and infers all four components of the Stokes' parameter. Using these observations, various parameters are inferred e.g., Line of Sight (LOS) magnetic field, Doppler velocities, vector magnetic fields etc. We have mainly used the LOS magnetogram for our analysis. These magnetograms have a 0.5{\arcsec} pixel size and {45} s cadence.

@@

The Interface Region Imaging Spectrograph \citep[IRIS;][]{iris} is a NASA small satellite explorer mission designed to observe the dynamics of the lower solar atmosphere. IRIS contains a spectrograph and a slit-jaw imager (SJI), observing the chromosphere, the transition region, and the lower corona. It primarily observes the Sun in two pass bands around 1400~{\AA} and 2800~{\AA}. IRIS provides data at high spatial resolution (0.33 {\arcsec} in FUV and 0.4 {\arcsec} in NUV), high time cadence (up to 1 second), and high spectral resolution (12 or 25 m{\AA}).

IRIS observes three wavelength bands: two in the Far Ultraviolet (FUV) range, spanning 1331.7{--}1358.4~{\AA} and 1389{--}1407~{\AA}, and one in the Near Ultraviolet (NUV) range, covering 2782.7{--}2851.1~{\AA}. IRIS measures spectra through two primary modes: raster and sit-and-stare. In raster mode, the slit traverses a field of view, capturing spectra from each pixel along its path. When the displacement between consecutive slit positions is comparable to the slit width, it is called dense raster mode; otherwise, it is called sparse raster mode. Alternatively, the instrument may position the slit on a specific region for observation of the Sun. This mode can allow the Sun to naturally rotate across the region or adjust for solar rotation while maintaining continuous observation. This mode is referred to as sit-and-stare. In our analysis, we have used both spectra and SJI imaging from the \ion{Mg}{2} h~\&~k window and SJI imaging from the \ion{Si}{4} 1402~{\AA} window.

@@

Solar Orbiter \citep[SO;][]{so} is a joint mission between the European Space Agency (ESA) and NASA, launched in February 2020. The Spectrometer/Telescope for Imaging X-rays (STIX) \citep[STIX,][]{stix,stix1} is designed to study the Sun's X-ray emissions, providing crucial insights into solar flares and other high-energy phenomena. STIX incorporates a set of collimators and detectors to capture high-resolution images of the Sun in X-rays. These images allow us to pinpoint the locations and intensities of X-ray emissions associated with solar flares. The instrument covers a broad range of X-ray energies, from approximately 4 to 150 keV, probing both thermal and non-thermal X-rays. STIX provides X-ray imaging with a spatial resolution $\sim$ 7 {\arcsec} at the shortest wavelengths using Fourier imaging techniques. Notably, the spatial resolution depends on the visibilities at various energies. STIX can operate at a temporal cadence as high as 0.1 s. The cadence of observation is adjustable depending on the flaring activity. The instrument operates in two data modes: the low-resolution background monitoring mode in quiet conditions and the high-resolution mode activated during solar flares. STIX works together with other instruments onboard the Solar Orbiter, providing complementary measurements to enhance our understanding of the Sun's dynamic behaviour across different wavelengths and energy ranges.

@@

Solar Terrestrial Relations Observatory-Ahead \citep[{\it STEREO-A};][]{stereo}, is one of two spacecraft in NASA's STEREO mission, which was launched in October 2006 with the goal of studying the Sun and its dynamic behaviour. {\it STEREO-A} and its twin spacecraft {\it STEREO-B} (Behind) were designed to provide stereoscopic views of the Sun, enabling three-dimensional observations of solar phenomena such as coronal mass ejections (CMEs) and solar flares. Extreme Ultraviolet Imager \citep[EUVI;][]{euvi} is one of the key instruments onboard the {\it STEREO-A} spacecraft. EUVI is designed to capture images of the Sun in the extreme ultraviolet (EUV) wavelength range. This range of wavelengths is particularly useful for studying the Sun's outer atmosphere, known as the corona, and observing features such as solar flares, coronal loops, and coronal holes.

EUVI can capture high-resolution images of the Sun's corona, with spatial resolutions in order of a few arcseconds when the respective spacecraft is at the perihelion. It observes the Sun in 171~{\AA}, 195~{\AA}, 284~{\AA} and 304~{\AA}, simultaneously. EUVI provides continuous observations of the Sun from the STEREO-A spacecraft's vantage point, located slightly ahead of Earth in its orbit around the Sun. This allows EUVI to monitor solar activity over extended periods and to track the evolution of solar features such as AR, flares, and CMEs. We have used 171~{\AA} and 195~{\AA} observations from EUVI for our analysis.

@@

{\it Hinode} \citep{hinode}, meaning "Sunrise" in Japanese, is a space mission launched by the Japan Aerospace Exploration Agency (JAXA) in collaboration with NASA and the UK Space Agency in September 2006. Also known as Solar-B before its launch, {\it Hinode} is dedicated to studying the Sun's magnetic field and its dynamic behaviour, focusing on understanding the mechanisms driving solar activity and space weather phenomena. The X-Ray Telescope \citep[XRT;][]{xrt} is one of the key instruments aboard the {\it Hinode} spacecraft.

XRT captures solar corona images in X-ray wavelengths, allowing us to study phenomena such as solar flares, coronal loops, etc. Its imaging capabilities enable us to investigate the solar corona's structure, dynamics, and heating mechanisms with a pixel size 1.028{\arcsec}/pix and a Field of View (FoV) size (34{\arcmin}$\times$34{\arcmin}). XRT utilizes narrow-band filters to isolate specific X-ray emission lines emitted by highly ionized elements in the solar corona sensitive to temperatures from 1 {--} 30 MK. XRT provides continuous observations of the solar corona from the vantage point of the {\it Hinode} spacecraft, which orbits the Earth in a Sun-synchronous polar orbit. The cadence of XRT observations can range from 30 {--} 60 s in the regular synoptic mode and upto 2 s in flare mode. In our analysis, we used XRT imaging observation from various filters. %\info{When you discussing all these instruments, you should have a scheme, i.e., what parameters you are going to be mentioning for each, e.g. spatial resolution, wavelength band, temperature coverage where available, spectral resolution etc... and be consistent throughout the chapter.}

%--------------------------------------------------------------
\section{Geostationary Operational Environmental Satellites (GOES)}
%--------------------------------------------------------------

Geostationary Operational Environmental Satellites (GOES) are a series of weather satellites operated by the National Oceanic and Atmospheric Administration (NOAA) in the United States. These satellites continuously monitor weather conditions across the Americas, including the United States, Canada, Mexico, and parts of South America. GOES data are used by meteorologists and weather forecasters to monitor and predict weather conditions, track severe weather events, and issue warnings and advisories to the public. The continuous monitoring provided by GOES satellites helps improve the accuracy and timeliness of weather forecasts. In addition to monitoring terrestrial weather, GOES satellites also monitor space weather conditions, including solar flares, coronal mass ejections (CMEs), and geomagnetic storms. This information is crucial for forecasting space weather events and assessing their potential impact on satellite communications, navigation systems, and power grids. Among the numerous instruments on-board {\it GOES}, we have used data from two instruments extensively {--} X-Ray Sensor (XRS) and Solar Ultra Violet Imager (SUVI).

@@

The X-Ray Sensor \citep[XRS;][]{xrs} is an instrument onboard {\it GOES}. The primary function of the XRS is to continuously monitor and measure full disk integrated solar soft X-ray emissions, particularly those associated with solar flares. XRS is designed to detect and measure solar X-ray emissions in two energy bands: short wavelength (0.5 to 4 {\AA}) and long wavelength (1 to 8 {\AA}) with a 2 s cadence during nominal operations.

@@

Solar Ultraviolet Imager \citep[SUVI;][]{suvi} is an instrument aboard {\it GOES}, designed to observe the Sun in the EUV spectrum, providing crucial data for monitoring solar activity and space weather. SUVI captures images of the Sun in 94 {\AA}, 131 {\AA}, 171 {\AA}, 195 {\AA}, 284 {\AA} and 304 {\AA} with a 2.5{\arcsec} pixel size. These wavelengths are very similar to AIA and provide complementary observations and very similar temperature coverage as AIA. The FoV of SUVI is rotated by $\sim~45^{o}$, and provides a larger radial distance coverage along the diagonals. In our analysis, we have used SUVI observations from 94 {\AA}, 131 {\AA}, 171 {\AA}, and 195 {\AA}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Techniques for Astrophysical Plasma Diagnostic}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We employ various techniques throughout this thesis to infer electron density, ion and electron temperature, velocity, and thermal structure of the plasma observed in both Chromosphere and Corona. It is standard practice to use atomic databases like CHIANTI \citep{chianti,chianti1} to derive the plasma properties. Below we describe the working principle of the methods used within the scope of this thesis.

@@

Differential emission measure analysis is a commonly employed method with instruments that observe multiple spectral bands to determine the temperature and emission measure of optically thin coronal plasma exhibiting multiple thermal components along a line of sight. However, the inversion process is often challenging due to its ill-posed nature and the frequent lack of sufficient constraints, making it an under-determined problem. One of the very first DEM calculation algorithm was {\it xrt\_dem\_iterative}, which was designed and validated for use with XRT data \citep{golub04,weber04}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Doppler Velocity from Spectra}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The Solar Ultraviolet Imaging Telescope \citep[SUIT;][]{ghosh16,article}, iss one of the seven payloads aboard the Aditya-L1 mission \citep{adityal1, aditya} led by the Indian Space Research Organization (ISRO). Launched on September 2, 2023, the satellite orbits around the Sun-Earth L1 point in a halo orbit. Equipped with eleven science filters, comprising eight narrow-band and three broad-band filters, SUIT possesses a distinctive capability to explore various heights within the solar photosphere and chromosphere. This capability aids in understanding the diverse physical processes involved in the transport of mass and energy across different layers of the Sun. Additionally, SUIT offers a unique opportunity to spatially measure the solar spectral irradiance in the Near Ultraviolet (NUV) range, a critical aspect in advancing our comprehension of Sun-climate relationships.

{\suit} is designed to deliver continuous full-disk and Region of Interest (RoI) solar images, boasting a plate scale of 0.7{\arcsec}. It possesses the ability to track RoIs while compensating for the effects of differential rotation, as outlined by \cite{suit_algo}. Additionally, SUIT is equipped with onboard intelligence to detect and localize flares, and it can autonomously adjust exposure times to prevent saturation. The primary scientific inquiries targeted by SUIT include \citep{suit_science, suit_main}:

\begin{itemize}
    \item Dynamic coupling between the lower and the middle solar atmosphere
    \item Measurement of solar spectral radiance in the NUV.
    \item Spectral energy distribution of solar flares in NUV
    \item Dynamics of chromospheric eruptive phenomena at various spatiotemporal scales
    \item Initiation of Coronal Mass Ejections(CMEs) and space weather: the kinematics of erupting prominence during the early phase.
\end{itemize}

@@

The Solar Ultraviolet Imaging Telescope \citep[SUIT;][]{ghosh16,article} is one of the seven payloads onboard the Aditya-L1 mission \citep{adityal1} of the Indian Space Research Organization (ISRO). With its 11 science filters (3 broadband and eight narrow band), SUIT will have the capability to probe different heights of solar atmosphere in photosphere and chromosphere to help us understand the various processes that transport mass and energy from one layer to another. Through SUIT imaging, we would be able to resolve solar flares spatially on the surface of the Sun, for the first time in near ultra-violet (NUV), which will help us to address the questions regarding their build-up and triggering mechanism. In addition, to measure the spatially resolved solar spectral irradiance within the wavelength range that is central for studying the Chemistry of oxygen and ozone in the Stratosphere of Earth's atmosphere.

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[width = \linewidth]{Figures/Krucker_2015_ApJ_802_19_page-0004.jpg}
    \caption[X-ray and optical imaging of the three flares at the peak of the flare.]{X-ray and optical imaging of the three flares at the peak time of the impulsive phase: the images are from HMI with the pre-flare image subtracted. The contours represent RHESSI Clean maps in the thermal (red, 12–15 keV) and non-thermal (blue, 30–80 keV) HXR range \citep{krucker15}.}
    \label{fig1}
\end{figure}

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[width = 0.7\linewidth]{Figures/nphys1741_page-0002.png}
    \caption[Averaged TSI variations during flares of various strengths.]{Averaged TSI variations during flares. The TSI time-series averages over four exclusive sets of solar flares of decreasing amplitude. The black and pink curves correspond respectively to the TSI measured by the PMOV6 and the DIARAD radiometers. The dashed lines correspond to the 95\% confidence levels, while the vertical line denotes the peak time of the flare \citep{kretzschmar10}.}
    \label{fig2}
\end{figure}
